---
title: "Adding a study"
author: "K.A. Stam"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: default
vignette: >
  %\VignetteIndexEntry{Adding a study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}\n"
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

myDir <- paste0(tempdir(), "/dummy_beFAIR")
dir.create(myDir, showWarnings = FALSE)
knitr::opts_knit$set(root.dir = myDir)
setwd(myDir)
```

```{r setup, warning=FALSE, error=FALSE, message=FALSE}
library(beFAIR)

## TODO
# comments from myself
# order should be:
# 1. Show addStudy function first
# 2. show necessary scripts
# 3. show how to change the template.yaml
```

<center>
## 🚧 _Under construction_ 🚧
</center>



## The Study pillar

In this vignette you will learn how to add a study to a FAIRground, how
to draft and fill-in the relevant codebooks that reside in the the **study**
pillar. This pillar is the first foundation of every FAIRground,
it holds the clinical trial or cohort study itself: who participated, when
events occurred, and how the study was designed. For example, Within the
`participant` table you can find variables like _age_ and _BMI_ and the
`studydesign` table can be used to look-up when a participant got immunized or
treated. For more information on the various codebooks, check the 
[Codebookreference](codebooks_reference.html).

To follow along, we'll first set up a dummy fairpoint project and explore the 
structure of a FAIRground. 

```{r, eval=FALSE, message=FALSE}
# Create a FAIRground
myDir <- paste0(tempdir(), "/dummy_beFAIR")
startProject(dir = myDir, dummy=TRUE)

# Inspect the created beFAIR
cat(normalizePath(myDir, dummy_beFAIR.proj), "\n")

# Open your newly generated project
rstudioapi::openProject(paste0(myDir, "/dummy_beFAIR.proj") newSession = TRUE)


```


Each study lives in its own sub-folder under
`studies/` and its data are described by a minimum of four standard codebooks,
`id_mapping`, `participant`, `studyarm` and `studydesign`

## Create a project and connect a database

Before adding a study, initialise a FAIRground directory with `startProject()`.

beFAIR stores all data in a SQL database. For this walkthrough we use
**SQLite** — no server required, just a single file on disk. Call
`FAIRgroundConnect()` with the RSQLite driver, then `databaseInitialize()` to
create the two core registry tables (`table_script` and `table_codebook`):

```{r, message=FALSE}
FAIRgroundConnect(drv = RSQLite::SQLite(), dsn = "fairpoint.sqlite")
# databaseInitialize()
```

> **In production:** replace the SQLite driver with your ODBC Data Source Name:
> ```r
> FAIRgroundConnect(dsn = "MY_PROJECT_DSN")
> ```

## `AddStudy()`

Use `addStudy()` to register a new study. Two arguments are required:

- `study_name` — unique human-readable name, used as the directory name
- `study_id` — exactly five alphanumeric characters, used as the prefix when generating anonymised `participant_id` values

Optionally pass a `registration_code` (ClinicalTrials.gov NCT number) to
auto-populate study metadata.

```{r}
# addStudy(study_name = "CoHSI1", study_id = "CHSI1")
```

## Folder structure

`addStudy()` creates the following layout under `studies/CoHSI1/`:

```{r}
# list.files("studies/CoHSI1", recursive = TRUE)
```

- **`meta.yaml`** — study-level metadata to fill in
- **`data/`** — YAML files pointing to your raw data sources
- **`scripts/`** — one auto-generated ETL script per codebook

## meta.yaml

The `meta.yaml` file is initialised with the values you passed to `addStudy()`.
Open it and fill in the remaining fields before running `databaseUpdate()`.

```{r, echo=FALSE}
# cat(paste(readLines("studies/CoHSI1/meta.yaml"), collapse = "\n"))
```

Fields to complete:

| Field | Description | Example |
|---|---|---|
| `study_organism` | Organism studied | `Homo sapiens` |
| `disease` | Disease under investigation | `malaria` |
| `pathogen` | Causative pathogen | `Plasmodium falciparum` |
| `registration_code` | ClinicalTrials.gov NCT number | `NCT02237586` |

## ETL scripts


| Codebook | Table | Description |
|---|---|---|
| `id_mapping` | id_mapping | Links each raw source identifier to an anonymised `participant_id` |
| `participant` | participant | One row per participant; demographics and enrollment metadata |
| `studyarm` | studyarm | One row per study arm; describes the intervention groups within a study|
| `studydesign` | studydesign | One row per event per study arm; the trial timeline |


Four ETL scripts are drafted automatically, one per codebook:

```{r}
# list.files("studies/CoHSI1/scripts")
```

Each script follows the same structure:

1. **Dependencies** — load the study `meta.yaml` and extract the `study_id`
2. **Load raw data** — read your source files (you write this)
3. **Data preparation** — clean and reshape (you write this)
4. **DATABOOK** — create the empty `databook` via `codebookToEntries()`
5. **DATA** — populate each field (you write this)
6. **Validate** — check the data against the codebook rules
7. **Clean environment** — keep only `databook`

The sections marked `!! write code !!` are the ones you need to complete.

---

### id_mapping.R

The `id_mapping` script is the **entry point** for every study. It maps each
raw **source identifier** from your clinical database to a stable, anonymised
**`participant_id`**. beFAIR generates the `participant_id` automatically
during `databaseUpdate()` by SHA-256 hashing `"<study_id>_<NNNN>"`
(zero-padded `participant_nr`) and taking the first six hex characters.

Your task is to produce a row per source identifier with:

- `participant_nr` — a sequential integer assigned to each unique individual
- `source_id` — the original identifier from the source system

Multiple `source_id` values can share the same `participant_nr` when a
participant appears under different IDs across data sources.

```{r, echo=FALSE, comment=''}
# writeLines(readLines("studies/CoHSI1/scripts/id_mapping.R"))
```

**Example — completed DATA section:**

```r
# participant_nr
databook$data$participant_nr <- as.integer(rawData$participant_nr)

# source_id
databook$data$source_id <- rawData$DonorID

# study_id
databook$data$study_id <- studyID
```

> **Note — reserved source IDs:** The values `"BLANK"`, `"STD"`, `"NEG_CTRL"`,
> and `"POS_CTRL"` are treated as control or blank sample labels and are
> **never** entered in `id_mapping`. When `checkSourceIdExist()` validates
> downstream tables (e.g. experiment or analysis data) it automatically skips
> these identifiers, so they will not raise a missing-ID warning even if they
> appear as `source_id` values in those scripts.

---

### participant.R

The `participant` script captures **one row per participant** with demographic
and enrolment information. The `participant_id` here must match the values
generated by the `id_mapping` script — this link is established automatically
by `databaseUpdate()`.

```{r, echo=FALSE, comment=''}
# writeLines(readLines("studies/CoHSI1/scripts/participant.R"))
```

Key fields to populate from your raw data:

| Field | Type | Example |
|---|---|---|
| `source_id` | character | `"VOL-001"` |
| `study_id` | character | `"CHSI1"` |
| `study_name` | character | `"CoHSI1"` |
| `studyarm_id` | character | `"CHSI1_vaccine"` |
| `age` | numeric | `32` |
| `sex` | character | `"M"` or `"F"` |
| `height` | numeric | `178` (cm) |
| `weight` | numeric | `75` (kg) |
| `reference_date` | character | `"2021-03-01"` |

**Example — completed DATA section:**

```r
# source_id
databook$data$source_id <- rawData$DonorID

# study_id
databook$data$study_id <- studyID

# study_name
databook$data$study_name <- metaInfo$name

# studyarm_id
databook$data$studyarm_id <- paste0(studyID, "_", rawData$Group)

# age
databook$data$age <- as.numeric(rawData$Age)

# sex
databook$data$sex <- ifelse(rawData$Gender == "Male", "M", "F")

# height
databook$data$height <- as.numeric(rawData$Height_cm)

# weight
databook$data$weight <- as.numeric(rawData$Weight_kg)

# reference_date
databook$data$reference_date <- format(as.Date(rawData$EnrolmentDate), "%Y-%m-%d")
```

---

### studyarm.R

The `studyarm` script describes the **intervention groups** in the study. Each
arm gets one row. The `studyarm_id` is the foreign key that links participants
(via `participant.studyarm_id`) to their arm.

Convention for `studyarm_id`: `<study_id>_<arm_label>` (e.g. `CHSI1_vaccine`).

```{r, echo=FALSE, comment=''}
# writeLines(readLines("studies/CoHSI1/scripts/studyarm.R"))
```

**Example — completed DATA section for a two-arm study:**

```r
# studyarm_id
databook$data$studyarm_id <- c("CHSI1_vaccine", "CHSI1_placebo")

# study_id
databook$data$study_id <- c(studyID, studyID)

# studyarm_name
databook$data$studyarm_name <- c("Vaccine group", "Placebo group")

# studyarm_description
databook$data$studyarm_description <- c(
  "Participants receiving PfSPZ Vaccine at 9e5 SPZ",
  "Participants receiving normal saline placebo"
)
```

---

### studydesign.R

The `studydesign` script encodes the **scheduled events in the trial timeline**,
with one row per event per arm. Together with `studyarm`, it gives a complete
picture of what happened to each group and when. Timepoints are expressed as
days relative to the reference date (day 0 = first intervention).

```{r, echo=FALSE, comment=''}
# writeLines(readLines("studies/CoHSI1/scripts/studydesign.R"))
```

**Example — completed DATA section for a vaccination + challenge trial:**

```r
arms  <- c("CHSI1_vaccine", "CHSI1_vaccine", "CHSI1_vaccine", "CHSI1_placebo")
n     <- length(arms)

# studyarm_id
databook$data$studyarm_id <- arms

# study_id
databook$data$study_id <- rep(studyID, n)

# timepoint (days from reference date)
databook$data$timepoint <- c(0L, 28L, 56L, 0L)

# event_type
databook$data$event_type <- c("vaccination", "vaccination", "challenge", "placebo")

# event_name
databook$data$event_name <- c("Prime", "Boost", "CHMI", "Placebo injection")

# product_name
databook$data$product_name <- c("PfSPZ Vaccine", "PfSPZ Vaccine", "PfSPZ Challenge", "Normal saline")

# dose
databook$data$dose <- c(9e5, 9e5, 3.2e3, NA_real_)

# dose_unit
databook$data$dose_unit <- c("SPZ", "SPZ", "SPZ", NA_character_)
```
