---
title: "Adding an analysis"
author: "K.A. Stam"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: default
vignette: >
  %\VignetteIndexEntry{Adding an analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Create temporary directory
dir <- paste0(tempdir(), "/beFAIR")
dir.create(dir, showWarnings = FALSE)
knitr::opts_knit$set(root.dir = dir)
setwd(dir)
```

```{r setup, warning=F, error=F, message=F}
library(basictabler)
library(dplyr)
library(htmltools)
library(beFAIR)
library(yaml)
```

<center>
## 🚧 _Under construction_ 🚧
</center>


## The Analysis Pillar

The **analysis** pillar is one of three data pillars in beFAIR. It stores 
data that represents some form of interpretation applied to study and/or 
experimental data — ranging from standard pre-processing pipelines to elaborate 
cluster analyses. Examples include fitting an ELISA dose-response curve, running
a Self-Organizing Map (SOM) clustering pipeline on CyTOF measurements, or 
applying a custom statistical model.

Because analyses may draw on samples from multiple studies and on the outputs of
multiple experiments, they are linked to the rest of the database via 
`experiment_id` and `study_id` references stored in their `meta.yaml`.

All analyses live inside the FAIRground directory. beFAIR directly
sources each ETL script, which must produce a `databook` object and load it
into the database. Analyses can declare explicit dependencies on upstream
database tables (e.g. `cytofpanel` from a specific experiment) via the
`dependencies` field in `meta.yaml`; these are tracked in `table_dependency`
and used to flag when upstream data has changed.

```{r, echo=F}
startProject(dir = dir)
list.files(dir)
```

```{r}
overviewFAIRground()
```

---

## Internal delivery

An analysis is added with `addAnalysis()`. You must supply a unique name and a
scientific type. We recommend following the naming convention
`[study]_[type]_[descriptor]`.

The supported analysis types are: `"SOM"`, `"OMIQ"`, `"ELISA"`, `"Array"`,
`"Luminex"`, and `"Glycoprofiling"`.

```{r}
addAnalysis("CoHSI1_SOM_surface", "SOM")
```

### The analysis folder

A new folder is created under `analyses/` named after the analysis:

```{r}
list.files("analyses/CoHSI1_SOM_surface")
```

It contains three components:

- **`meta.yaml`**: metadata about the analysis, pre-filled from the base
  `analysis` codebook and the type-specific codebook (here `analysis_som`).
- **`data/`**: for storing data location descriptors (YAML files), same
  pattern as studies and experiments.
- **`scripts/`**: ETL R scripts drafted automatically for the chosen type.

### meta.yaml

The `meta.yaml` stores essential metadata about the analysis. beFAIR
pre-fills fields from the `analysis` codebook and any type-specific extension
(e.g. `analysis_som` adds `pipeline_location` and `pipeline_description`).
Fill in the missing values before running `databaseUpdate()`.

```{r, echo=F}
cat(paste(readLines("analyses/CoHSI1_SOM_surface/meta.yaml"), collapse = "\n"))
```

Key fields to complete:

| Field | Description |
|---|---|
| `study_ids` | Semicolon-separated IDs of studies whose samples were used |
| `experiment_ids` | Semicolon-separated IDs of source experiments |
| `description` | Free-text description of the analysis |
| `pipeline_location` | *(SOM only)* Path or URL to the SOM pipeline |

The `dependencies` field is initially empty. Fill it in to declare which
upstream database tables this analysis uses (e.g.
`- cytofpanel[Eef536a]`). beFAIR will track these in `table_dependency`
and flag the analysis as out-of-date when upstream data changes.

### ETL scripts

For a SOM analysis, two ETL scripts are drafted:

```{r}
list.files("analyses/CoHSI1_SOM_surface/scripts")
```

- **`som_clusterphenotype.R`**: loads the cluster phenotype table (one row per
  SOM cluster, with marker expression profiles).
- **`som_measurement.R`**: loads per-sample SOM measurements (cluster
  frequencies or similar metrics).

Each script follows the standard beFAIR ETL pattern — it must produce a
`databook` object and end with
`rm(list = setdiff(ls(), c("databook")))`. See `?draftScript` for details.


