---
title: "Adding an experiment"
output:
  rmarkdown::html_vignette: default
vignette: >
  %\VignetteIndexEntry{Adding an experiment} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}\n"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Create temporary directory
dir <- paste0(tempdir(), "/beFAIR")
dir.create(dir)
knitr::opts_knit$set(
  root.dir = dir
)

setwd(dir)

```

```{r setup, warning=F, error=F, message=F}
library(basictabler)
library(dplyr)
library(htmltools)
library(beFAIR)
library(yaml)
```

<center>
## 🚧 _Under construction_ 🚧
</center>


## The Experimental pillar

One of three data pillars is the `experiments` pillar. These contain the experimental 
data, typically generated in the lab from samples acquired from volunteers who
participated in the clinical trials (or studies). It's important to note that 
experiments may have been performed on samples from different studies. Therefore,
the data pillars are interconnected through the `event_id`. Any additional information
required to better understand the experimental data can be retrieved from the 
`participant` or `event` table. To get an quick overview of the folder structure
of you FAIRPoint you can run `overviewFAIRground()` 


```{r, echo=F}
## TODO I want a `full` demo beFAIR so we have more to show in the vignette
dir <- paste0(tempdir(), "/beFAIR")
startProject(dir = dir)

list.files(dir)
```

```{r}
overviewFAIRground()
```

A new experiment can be added by running `addExperiment()`. A unique study
name together with the type of experiment is required and will kickstart the 
necessary ETL scripts under `/experiments/[experiment_name]`. Any unique name is
possible, but we strongly recommend using our naming convention: [reference to studies]_
[experiment type]_[other specifics]. The experiment type is there to ask for 
extra relevant meta information, e.g. machine specifics, but will also kickstart
technique specific ETL scripts, e.g. for cytof we need a panel that specefies 
which antibodies and metals were used for the experiment. It's possible to
add an experiment without a experiment_type, but it is not recommended. 

## The experiment folder

```{r}
addExperiment(experiment_name = "CoHSI1_CyTOF_exvivo",
              experiment_type = "CyTOF")
```

A new folder can be found in the `experiments` directory, named after the experiment name you provided (e.g., `CoHSI1_CyTOF_exvivo`).

In this folder, you will find the following components:

- **`meta.yaml`**: This file contains essential metadata about the experiment
- **`data/`**: A folder designated for storing the data locations.
- **`scripts/`**: A folder with the ETL scripts related to the experiment

We will break down each of these components in the following sections for a clearer understanding.

### meta.yaml

Just as with studies, you will find a `meta.yaml` file located at `/experiments/[experiment_name]/meta.yaml`. 
This file contains essential metadata about the experiment, which you are 
requested to supplement with the missing information. The table below is a 
small snapshot of information stored in the meta.yaml file.

```{r, echo=F}
studyMeta <- yaml::read_yaml("experiments/CoHSI1_CyTOF_exvivo/meta.yaml")
tbl <- BasicTable$new()
tbl$addData(data.frame(studyMeta)[,2:10])
tbl$renderTable()
```

### Data folder

The data folder in each study directory is intended to organize and manage the locations of the raw data files associated with the study. This is facilitated by YAML files, where each raw data file can have its own YAML configuration file. These YAML files provide key metadata about the raw data, including its location, file type, and contact information of the data owner.

This information is used to automate the extraction and transformation processes within the study's ETL pipeline. The YAML files ensure that raw data is easily located and interpreted in a structured, standardized format. 

An example YAML file,  `template.yaml`, is automatically placed in the data folder of a new study. You can edit the template.yaml file or create additional YAML files for each raw data file in your study. Below is the structure of the template:

```{r, echo=F}
yaml_content <- readLines(paste0("experiments/CoHSI1_CyTOF_exvivo/data/template.yaml"))
cat("```yaml\n", paste(yaml_content, collapse = "\n"), "\n```\n")
```

### ETL scripts

In the scripts folder, the ETL (Extract, Transform, Load) scripts are provided in R format. These scripts correspond to different entities: participants, events, and samples. Each script follows the data structure defined in the codebook and is responsible for extracting, transforming, and validating the data before it can be loaded into the data warehouse.

```{r}
list.files("experiments/CoHSI1_CyTOF_exvivo/scripts")
```

