---
title: "Codebook reference"
author: "K.A. Stam; F. de Weijer"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Codebook reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(yaml)
```

This page documents all codebooks included in beFAIR. Each codebook defines the schema for a database table — its variables, types, and validation rules.

See [Codebook configuration guide](codebook_configuration.html) for details on the YAML format.

```{r helpers, include=FALSE}
desc_file <- system.file("codebooks/codebook_description.yaml", package = "beFAIR")
table_descriptions <- yaml::read_yaml(desc_file)

safe_str <- function(x) {
  if (is.null(x) || (length(x) == 1 && is.na(x))) "" else trimws(as.character(x))
}

render_codebook_section <- function(cb_name, cb_path) {
  cb <- yaml::read_yaml(cb_path)

  td <- table_descriptions[[cb_name]]
  description  <- if (!is.null(td$description))  trimws(td$description)  else NULL
  primary_key  <- if (!is.null(td$primary_key)  && length(td$primary_key)  > 0)
    paste(sapply(td$primary_key,  as.character), collapse = ", ") else NULL
  foreign_keys <- if (!is.null(td$foreign_keys) && length(td$foreign_keys) > 0)
    paste(sapply(td$foreign_keys, as.character), collapse = ", ") else NULL

  cat(sprintf("\n### `%s`\n\n", cb_name))

  if (!is.null(description)) cat(description, "\n\n")

  meta <- sprintf("**Version:** %s", safe_str(cb$version))
  if (!is.null(primary_key))
    meta <- paste0(meta, " &nbsp;&nbsp;|&nbsp;&nbsp; **Primary key:** `", primary_key, "`")
  if (!is.null(foreign_keys))
    meta <- paste0(meta, " &nbsp;&nbsp;|&nbsp;&nbsp; **Foreign keys:** `", foreign_keys, "`")
  cat(meta, "\n\n")

  if (is.null(cb$variables) || length(cb$variables) == 0) {
    cat("_No variables defined._\n\n")
    return(invisible(NULL))
  }

  rows <- lapply(names(cb$variables), function(var_name) {
    v <- cb$variables[[var_name]]
    role <- if (isTRUE(v$primary_key)) {
      "PK"
    } else if (!is.null(v$foreign_key)) {
      paste0("FK \u2192 ", v$foreign_key)
    } else {
      ""
    }
    data.frame(
      Variable    = var_name,
      Role        = role,
      Description = safe_str(v$description),
      Type        = safe_str(v$type),
      `SQL Type`  = safe_str(v$type_sql),
      Categories  = safe_str(v$categories),
      Labels      = safe_str(v$labels),
      stringsAsFactors = FALSE,
      check.names      = FALSE
    )
  })
  df <- do.call(rbind, rows)
  print(knitr::kable(df, format = "html", escape = TRUE))
  cat("\n")
}

render_pillar <- function(pillar) {
  dir_path <- system.file(paste0("codebooks/", pillar), package = "beFAIR")
  files    <- sort(list.files(dir_path, pattern = "\\.yaml$", full.names = TRUE))
  for (f in files) {
    cb_name <- tools::file_path_sans_ext(basename(f))
    render_codebook_section(cb_name, f)
  }
}
```

## Study

Tables related to participants, events, and study design.

```{r study, results='asis'}
render_pillar("study")
```

## Experiment

Tables related to laboratory experiments and measurement panels.

```{r experiment, results='asis'}
render_pillar("experiment")
```

## Analyses

Tables containing processed measurement results.

```{r analyses, results='asis'}
render_pillar("analyses")
```

## Common

Internal registry tables used by beFAIR to track scripts and codebook versions.

```{r common, results='asis'}
render_pillar("common")
```
