---
title: "Database: Smoke test"
author: "K.A. Stam"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: default
vignette: >
  %\VignetteIndexEntry{Database: Smoke test} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}\n"
---

```{r setup, include=FALSE}
library(beFAIR)
knitr::opts_chunk$set(echo = TRUE)
```

<center>
## 🚧 _Under construction_ 🚧
</center>


## Setting up a database

After creating a FAIRground 
([Starting a project](starting_a_project.html)), the next step is to set up a
telational database. The `beFAIR` package is designed around interaction
with a SQL database. All database operations are implemented via the `DBI` 
interface, which allows beFAIR to work with different backends 
(e.g., SQLite, PostgreSQL, MariaDB).

How and where you host your database depends on your institutional 
infrastructure. If you do not yet have a database configured, you can still 
follow this vignette to understand how beFAIR manages database interactions.

## Connection

The most straightforward way to connect to a database is via a DSN 
(Data Source Name) connection. On Windows, DSNs can be configured using the 
__ODBC Data Sources__ administrator.

Once configured, you can connect using `FAIRgroundConnect()`. This establishes a 
default connection that will be used throughout your session.

```{r, eval = FALSE}
## How you normally would connect to the database 
FAIRgroundConnect("LUCID_beFAIR")
con <- FAIRgroundConnection() # If you want to retrieve the active connection
```

The active connection is stored internally and reused by beFAIR functions 
unless explicitly overridden.

To ensure reproducibility, this vignette uses an in-memory SQLite database via 
`RSQLite`. We also initialize a dummy FAIRground containing example data 
and ETL scripts.

```{r}
con <- FAIRgroundConnect(drv = RSQLite::SQLite(), dsn = ":memory:")

dir <- file.path(tempdir(), "/beFAIR")
startProject(dir = dir, dummy=TRUE)

## Location of your dummy beFAIR
# cat(normalizePath(dir))
```


```{r}
databaseInitialize()
DBI::dbReadTable(con, "table_script")
```

