---
title: "Developer internals: databaseUpdate() and the registry tables"
author: "K.A. Stam"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Developer internals: databaseUpdate() and the registry tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

```{r insert-icons, echo=FALSE, eval=TRUE}
find_img <- function(name) {
  candidates <- c(file.path("images", name),
                  file.path("vignettes", "images", name))
  candidates[file.exists(candidates)][1]
}
uri_codebook  <- knitr::image_uri(find_img("icon_codebook.svg"))
uri_etlscript <- knitr::image_uri(find_img("icon_ETLscript.svg"))

htmltools::tags$script(htmltools::HTML(sprintf('
document.addEventListener("DOMContentLoaded", function() {
  var iconMap = {
    "table_codebook": "%s",
    "table_script":   "%s"
  };
  document.querySelectorAll("h3").forEach(function(h3) {
    var code = h3.querySelector("code");
    if (code && iconMap[code.textContent]) {
      var img = document.createElement("img");
      img.src = iconMap[code.textContent];
      img.width = 35;
      img.style.verticalAlign = "middle";
      img.style.marginRight = "0.4rem";
      h3.insertBefore(img, h3.firstChild);
    }
  });
});
', uri_codebook, uri_etlscript)))
```

This guide explains how beFAIR manages its database internally. It covers
the two registry tables that track schema state and ETL script history, and
walks through every function called during `databaseUpdate()`. Read this if
you need to understand, debug, or extend the system.

---

## 1. The registry tables

beFAIR uses two permanent registry tables to coordinate all database
operations. They are created once by `databaseInitialize()` and are never
dropped or truncated by the ETL cycle.

### `table_codebook`

One row per codebook in the package. Tracks the canonical schema definition
and is the reference point for version comparisons.

| Column | SQL type | Purpose |
|---|---|---|
| `codebook_id` | VARCHAR | Relative path key, e.g. `study/participant`; custom codebooks use `custom/study/my_assay` |
| `codebook_name` | VARCHAR | Matching database table name, e.g. `participant` |
| `codebook_version` | VARCHAR | SemVer-style string, e.g. `"2.1"` |
| `authors` | VARCHAR | Codebook author(s) |
| `created_at` | VARCHAR | ISO-8601 creation date |
| `updated_at` | VARCHAR | ISO-8601 last-modified date |
| `n_parameters` | INT | Number of variables in the codebook |
| `codebook_origin` | VARCHAR | `"beFAIR"` for built-in package codebooks; the project directory basename (e.g. `"MyStudy2024"`) for custom codebooks |

**Version semantics.** Codebook versions are interpreted as `major.minor`
numeric pairs:

- **Major bump** (e.g. `1.x → 2.x`): the associated data table is **dropped
  and recreated**; all data in that table is lost. Use this only when the
  schema is incompatible with existing data.
- **Minor bump** (e.g. `1.1 → 1.2`): only the `table_codebook` metadata row
  is updated; the data table and its contents are untouched. Use this for
  description edits, adding optional fields, or other non-breaking changes.
  Note that all associated ETL scripts will be **queued for a re-run.**

### `table_script`

One row per script registered with the system. Tracks what was run, when,
and whether the output was valid. This includes some flags for *TODO's* and
*warnings* in the ETL script.

| Column | SQL type | Purpose |
|---|---|---|
| `script_id` | VARCHAR | Relative file path, e.g. `studies/CHHIL/scripts/participant.R`; for external delivery this is the retrieval script, e.g. `analyses/myAnalysis/scripts/retrieve_myAnalysis.R` |
| `script_hash` | VARCHAR | MD5 of the script file; **change triggers re-run** |
| `codebook_name` | VARCHAR | Target table for this script; `NA` for external retrieval scripts (they load multiple tables) |
| `codebook_version` | VARCHAR | Codebook version at last run; **drift triggers re-run** (internal scripts only) |
| `authors` | VARCHAR | Author(s) from script metadata |
| `created_at` | VARCHAR | Date first added to the registry |
| `updated_at` | VARCHAR | Date of last successful run |
| `run_time` | DECIMAL | Execution time in seconds |
| `flag_TODO` | INT | Count of `#TODO` comments found in the script |
| `flag_WARN` | INT | Count of warnings captured during execution |

### Querying the registry tables

```{r query-registry}
FAIRgroundConnect("LUCID_FAIRPOINT")
con <- FAIRgroundConnection()

DBI::dbReadTable(con, "table_codebook")
DBI::dbReadTable(con, "table_script")
```

You can also run these during a live session to inspect state mid-update when
debugging.

---

## 2. `databaseUpdate()` — the orchestrator

`databaseUpdate()` coordinates a complete ETL cycle.
Its high-level call sequence is:

<pre>
databaseUpdate()
├── FAIRgroundConnection()           # resolve default connection
├── <a href="#single-user-lock">setSingleUser()</a>                 # lock DB to exclusive mode (MSSQL only)
├── on.exit:
│   ├── <a href="#on-exit-guarantee">setMultiUser()</a>              # restore multi-user mode
│   └── <a href="#on-exit-guarantee">FAIRgroundDisconnect()</a>       # close connection
├── <a href="#prepare-codebooks">databasePrepareCodebooks()</a>      # sync codebook schema → DB
├── <a href="#prepare-scripts">databasePrepareScripts()</a>        # discover scripts needing (re)run
├── for each script:
│   ├── <a href="#remove-entries">sqlRemoveEntries()</a>          # delete old rows written by this script
│   ├── <a href="#source-script">sourceScript()</a>              # execute ETL script, get databook
│   ├── <a href="#validation-gate">[validation gate]</a>           # checks databook$validation$status
│   └── <a href="#prepare-data">databasePrepareData()</a>       # upload data & update table_script
├── <a href="#post-loop">databaseUpdateOverview()</a>        # rebuilds study, experiment and analysis table
└── <a href="#post-loop">databaseUpdatePublications()</a>    # rebuild publications table
</pre>

### Single-user lock (MSSQL) {#single-user-lock}

When connecting to a Microsoft SQL Server database, `databaseUpdate()` calls
`setSingleUser()` immediately after establishing the connection. This
acquires an exclusive lock on the database, preventing concurrent writes
from other sessions during the ETL cycle. The lock is always released by
`setMultiUser()` in an `on.exit()` handler — even if the function stops
early due to an error.

SQLite connections do not support single-user mode; `setSingleUser()` and
`setMultiUser()` are no-ops for that backend.

### `on.exit()` cleanup guarantee {#on-exit-guarantee}

The `on.exit()` block (with `add = TRUE`) runs unconditionally when
`databaseUpdate()` exits, regardless of whether it completes normally or
throws an error. It restores multi-user mode and closes the connection.
This means you never need to call `FAIRgroundDisconnect()` manually after
`databaseUpdate()`.

### Logging

Pass `enable_logging = TRUE` to write a JSON log for the entire update cycle:

```{r enable-logging}
FAIRgroundConnect("LUCID_FAIRPOINT")
databaseUpdate(enable_logging = TRUE)
```

Log files are written to `logs/files/<timestamp>.json` (e.g.
`logs/files/2024-03-15_09-30-00.json`). Each log entry is a JSON object with
the severity level, timestamp, message, and any named fields passed by the
caller.

To read a log after a run:

```{r read-log}
library(jsonlite)

log_file <- tail(list.files("logs/files", full.names = TRUE), 1)
logs <- stream_in(file(log_file))
```

See [Section 5](#logging) for more detail on the logger setup.

---

## 3. Step-by-step breakdown

### 3.1 `databasePrepareCodebooks()` {#prepare-codebooks}

Synchronizes the YAML codebook definitions in the package with `table_codebook`
and the actual data tables in the database.

**Steps:**

1. Read current `table_codebook` from the database.
2. **Migration check**: if the `codebook_origin` column is absent (existing
   databases pre-v0.65), execute:
   ```sql
   ALTER TABLE table_codebook ADD COLUMN codebook_origin VARCHAR(255);
   UPDATE table_codebook SET codebook_origin = 'beFAIR';
   ```
3. Build the expected codebook table from package YAMLs **and** any custom
   codebooks in `codebooks/` via `codebookToTable()`. Each row carries a
   `codebook_origin` value: `"beFAIR"` for built-in codebooks, or the
   project directory basename for custom ones.
5. Loop over each codebook entry and classify it:

```
For each codebook (package or custom):
│
├── Not in table_codebook?
│   └── NEW → sqlGenerateTable() + sqlPushTable() into table_codebook
│             (row includes codebook_origin)
│
└── Already in table_codebook:
    ├── major version increased?
    │   └── MAJOR BUMP → DROP TABLE + sqlGenerateTable()
    │                  → DELETE old row + sqlPushTable() new row
    │                  ⚠ All data in this table is lost
    │
    ├── minor version increased?
    │   └── MINOR BUMP → UPDATE table_codebook metadata only
    │                   (version, authors, updated_at, n_parameters)
    │
    └── version unchanged?
        └── No action
```

All detected changes are logged via the beFAIR logger.

### 3.2 `databasePrepareScripts()` {#prepare-scripts}

Discovers which scripts need to be (re)run and whether each uses internal
or external delivery.

**Steps:**

1. Read `table_script` from the database.
2. Read `table_codebook` to get current codebook versions.
3. Scan `studies/*/scripts/`, `experiments/*/scripts/`, and
   `analyses/*/scripts/` for `.R` files.
4. Compute the MD5 hash of the script file.
5. Classify each script:
   - **New**: path not in `table_script` → insert a placeholder row. Only
     inserted if the inferred `codebook_name` (from the filename) exists in
     `table_codebook`; otherwise log an error and skip.
   - **Hash changed**: file was edited since the last run.
   - **Codebook version drifted**: `table_script.codebook_version` is behind
     `table_codebook.codebook_version`.
   - **Up to date**: no action.
6. **Suspended scripts**: if a filename starts with `_` (e.g.
   `_participant.R`), the script is skipped entirely. The function checks
   whether the corresponding `script_id` (without the underscore) has
   existing rows in the database and logs a warning if so. This lets
   maintainers safely disable a script without losing the audit trail.
7. Return a data frame of scripts to run with columns `script_id`,
   `codebook_name`, and `script_hash`, ordered by dependency:
   `id_mapping → studyarm → studydesign → participant → event`, then
   experiments, then analyses.

### 3.3 Per-script loop

`databaseUpdate()` iterates over the data frame returned by
`databasePrepareScripts()` and runs the standard four-step ETL flow for
each script.

---

##### 3.3.1 `sqlRemoveEntries()` {#remove-entries}

Before re-running a script, its previously uploaded rows are deleted:

```sql
DELETE FROM <codebook_name> WHERE script_id = '<script_id>'
```

This ensures that the re-run starts with a clean slate and there are no
duplicate or stale rows from a previous version of the script.

##### 3.3.2 `sourceScript()` {#source-script}

Executes the ETL script and returns a `databook` object.

- The script runs in an **isolated environment** (`new.env()`) so it cannot
  accidentally read or overwrite variables in the calling session.
- The active DBI connection (`con`) is injected into this environment as the
  only external dependency.
- Before execution, all `#TODO` comments in the source text are counted
  (`flag_TODO`).
- During execution, R warnings are intercepted and counted (`flag_WARN`);
  messages are muffled to suppress `pointblank` output.
- Execution time is measured with `Sys.time()`.
- If the script throws an error, `sourceScript()` logs it and returns `NULL`;
  the main loop skips to the next script.

**The `databook` contract.** Every ETL script must produce a named list
called `databook` with these fields:

| Field | Content |
|---|---|
| `$data` | A data frame with the rows to upload |
| `$codebook` | The full codebook list object |
| `$codebook_name` | Character; target table name |
| `$codebook_version` | Character; version string |
| `$validation` | Output of `validateDataByCodebook()` |
| `$run_time` | Numeric; execution time in seconds (added by `sourceScript()`) |
| `$flag_TODO` | Integer; `#TODO` count (added by `sourceScript()`) |
| `$flag_WARN` | Integer; warning count (added by `sourceScript()`) |

`codebookToEntries(cb_name, nrow = N)` creates the correctly-typed empty
`$data` skeleton. Scripts should build their data in this frame to guarantee
column types match the database schema.

A minimal ETL script skeleton looks like this:

```{r etl-skeleton}
# --- Setup ---
databook <- codebookToEntries("participant", nrow = 100)

# --- Fill data ---
databook$data$source_id   <- raw_data$subject_id
databook$data$study_id    <- "MYST1"
databook$data$age         <- raw_data$age_years
# ...

# --- Validate ---
databook$validation <- validateDataByCodebook(databook)

# --- Clean up (required) ---
rm(list = setdiff(ls(), c("databook")))
```

The final `rm()` call is required. `sourceScript()` reads `databook` from
the script's local environment after `source()` returns; leaving other
variables behind does not cause errors but wastes memory.

##### 3.3.3 Validation gate {#validation-gate}

`databaseUpdate()` checks `databook$validation$status`:

- `1` (pass) → proceed to `databasePrepareData()`
- `0` (fail) → log the failure, skip upload, continue with the next script

`validateDataByCodebook()` uses the `pointblank` package to apply the
validation rules in the codebook YAML to `databook$data`. It also calls
`checkSourceIdExist()` for any table that has a `source_id` column (except
`id_mapping` itself), which confirms that every non-reserved `source_id`
value exists in the `id_mapping` table.

**Reserved `source_id` values** — `BLANK`, `STD`, `NEG_CTRL`, `POS_CTRL`,
`CTRL`, `BLANK_STD`, `SAMPLE_STD` — bypass the `id_mapping` lookup. These
represent control and blank samples that do not correspond to real
participants.

##### 3.3.4 `databasePrepareData()` {#prepare-data}

Transforms the validated data and uploads it to the database, then updates
`table_script`.

**For `id_mapping` scripts:**

1. Call `addParticipantIds()` to generate a deterministic `participant_id`
   for each row. The ID is computed as the first 6 hex characters of the
   SHA-256 hash of the string `"<study_id>_<NNNN>"` (4-digit zero-padded
   `participant_nr`), yielding a value like `"MYST1_a3f9b2"`.
2. Add the `script_id` column.
3. Upload via `sqlPushTable()`.

**For all other tables:**

1. If `source_id` is present: left-join against the `id_mapping` table to
   attach `participant_id`. Reserved IDs receive their own value as
   `participant_id` (e.g. `participant_id = "BLANK"`).
2. If both `event_id` and `timepoint` are present: construct
   `event_id = paste0(participant_id, "_", sprintf("%04d", timepoint))`.
   Reserved IDs get `event_id = NA`.
3. Add the `script_id` column.
4. Upload via `sqlPushTable()`.

After a successful upload, `updateTableScript()` updates the `table_script`
row for this script with the current MD5 hash, codebook version,
`updated_at` timestamp, `run_time`, `flag_TODO`, and `flag_WARN`.

### 3.4 Post-loop: `databaseUpdateOverview()` and `databaseUpdatePublications()` {#post-loop}

After all scripts have been processed, two aggregate tables are always
rebuilt unconditionally — even if no individual scripts were updated:

- `databaseUpdateOverview()` rebuilds the `study`, `experiment`, and
  `analysis` aggregate tables.
- `databaseUpdatePublications()` rebuilds the `publications` table.

Because these always run, they reflect the current state of the database
regardless of which scripts ran in this cycle.

---

## 4. Supporting SQL helpers

### `sqlGenerateTable(cb_name)`

Reads the codebook YAML for `cb_name`, builds a `CREATE TABLE` DDL string
from the `type_sql` fields of each variable, and executes it. `DECIMAL`
columns are always created as `DECIMAL(38, 18)` to preserve precision.

Example of the generated DDL for a simplified codebook:

```sql
CREATE TABLE participant (
  script_id VARCHAR,
  source_id VARCHAR,
  study_id VARCHAR,
  participant_nr INT,
  participant_id VARCHAR,
  age DECIMAL(38, 18),
  sex VARCHAR
);
```

### `sqlPushTable(data, table_name, script_id)`

Wraps `DBI::dbWriteTable(..., append = TRUE)`. On success, logs the number
of rows written. On failure, logs the error message and returns `FALSE`.

### `sqlRemoveEntries()`

Issues a parameterized `DELETE` statement for the given `codebook_name` and
`script_id`. See [Section 3.3.1](#331-sqlremoveentries) above.

---

## 5. Logging {#logging}

beFAIR uses the `lgr` package with a named logger hierarchy. All internal
functions obtain the logger via:

```{r get-logger}
lg <- lgr::get_logger("beFAIR")
```

This creates a child logger under the root `lgr` logger, allowing the
threshold and appenders to be configured independently.

### `.initLogger()` and `.disableLogger()`

`.initLogger()` is called by `databaseUpdate()` when `enable_logging = TRUE`:

- Clears any existing appenders on the `"beFAIR"` logger.
- Sets the threshold to `"info"`.
- Adds a `lgr::AppenderFile` with `lgr::LayoutJson` that writes to
  `logs/files/<YYYY-MM-DD_HH-MM-SS>.json`.

`.disableLogger()` removes all appenders (flushing pending writes to disk)
and is called in the `on.exit()` handler at the end of `databaseUpdate()`.

### Reading log files

```{r read-logs}
library(jsonlite)

# Find the most recent log file
log_file <- tail(list.files("logs/files", full.names = TRUE, pattern = "\\.json$"), 1)

# Stream-parse the JSON log (one object per line)
logs <- jsonlite::stream_in(file(log_file), verbose = FALSE)

# Filter to warnings and errors only
logs[logs$level >= 300, c("timestamp", "level_name", "msg", "script_id")]
```

Log levels follow the `lgr` convention: `info = 200`, `warn = 300`,
`error = 400`, `fatal = 500`.
